Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-08-07 doc/invt.tex
modify invt.tex

\documentclass[preprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of flat-distribution sampling methods
  for free energy calculation including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the error in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme.
  %
  While the asymptotically optimal schedule for
  the single-bin updating scheme (commonly used in WL)
  is given by an inverse-time formula,
  that for the Gaussian updating scheme (commonly used in metadynamics)
  is often more complex.
  %and the initial updating magnitude is ideally half of
  %the previous equilibrium value.
  %
  Further,
  we show that the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  These bandpass updating schemes aim at
  partially flattening a few selected histogram modes
  and their optimal schedule
  is also given by the inverse-time prescription.
  %
  Constructed from orthogonal polynomials,
  bandpass updating schemes serve as generalizations
  of the Langfeld-Lucini-Rago algorithm
  and can be used for parameter tuning for the method of Gaussian ensemble
  and umbrella sampling.
  %
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian versions.
%
We will map the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
Further, we will show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter allow one to partially flatten
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.
%
The optimal schedule of bandpass updating schemes
is also an inverse-time schedule.
%
Constructed from orthogonal polynomials,
bandpass updating schemes serve as
generalizations of the Langfeld-Lucini-Rago (LLR)
algorithm\cite{langfeld2012, pellegrini2014},
and can help other FDS
algorithms\cite{neuhaus2006, *neuhaus2007, zhu2012}
adjust the parameters of the bias potential automatically.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we briefly review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we derive the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

From Eqs.
\eqref{eq:vt_diffeq} and \eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  %=
  %\alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot \vartheta\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
\vartheta(q')
&\equiv
e^{q' - q(T)}.
\label{eq:theta_def}
\end{align}

@@

The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
By a proper overall scaling of $\mathbf w$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will %take advantage of the analogy and
borrow techniques used
in studying transition matrices\cite{vankampen},
but limit ourselves to properties that are unaffected
by the admission of negative elements into $\mathbf w$.

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $x_i$,
%
\begin{equation}
  {\tilde x}_k
  \equiv \mathcal{F}[x_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, x_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
$x_i = \sum_{k = 0}^{n-1} \phi_{ki} \, \tilde{x}_k / \rho_i$.
%
%\begin{equation}
%  x_i = \frac{1}{\rho_i} \sum_{k = 0}^{n-1} \phi_{ki} \, \tilde{x}_k
%  .
%  \notag
%  %\label{eq:gft_inv}
%\end{equation}
%
Then,
we can rewrite Eq. \eqref{eq:ortho0} as
$\mathcal F[1]_k = \delta_{k0}$,
%
and Eq. \eqref{eq:eigenmode0} means that
the bin average
%of any quantity, $x_i$, is the first mode of the Fourier transform:
$\bav{x}= \mathcal F[x_i]_0 = \tilde x_0$,
%
and
\begin{equation}
  \tilde x_{*k}
  = \tilde x_k - \bav{x} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0
    ,
    \\
    \tilde x_k  & k > 0
    .
  \end{dcases}
  \label{eq:xtstar}
\end{equation}
%
Thus, the only effect
of the bin-average deduction on $\tilde x_k$ is to
annihilate the $k=0$ mode.
%
We also have Parseval's theorem,
\begin{equation}
  \sum_{k=1}^{n-1} \tilde x_k \, \tilde y_k
  =
  \sum_{k=0}^{n-1} \tilde x_{*k} \, \tilde y_{*k}
  =
  \sum_{i=1}^n x_{*i} \, y_{*i}
  \label{eq:parseval}
\end{equation}

@@

By Eq. \eqref{eq:parseval},
we can rewrite the error defined in Eq. \eqref{eq:error_def} as
$\Err(T) = \sum_{k = 1}^{n - 1} \left\langle {\tilde v}_k^2(T) \right\rangle$.
%
%\begin{align}
%  \Err(T)
%  =
%  \sum_{k = 1}^{n - 1}
%    \left\langle
%      {\tilde v}_k^2(T)
%    \right\rangle
%  .
%\notag
%\end{align}
%
As in Eq. \eqref{eq:error_tot},
we can decompose the error as the sum
of the residual error and the asymptotic error, with
\begin{align}
  \Err_R(T)
  &=
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) }
  ,
  \label{eq:error_res}
  \\
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot \vartheta_k^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  \vartheta_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\bigl\langle {\tilde \zeta}_k(0)
\, {\tilde \zeta}_k(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty
  \bigl\langle {\tilde \zeta}_k(0)
            \, {\tilde \zeta}_k(t) \bigr\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum} represents
a first approximation for modeling
a general histogram fluctuation, ${\tilde \zeta}_k(t)$,
as an equivalent white noise.
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma_measure},
we discuss a way of estimating $\Gamma_k$
from an FDS simulation under a constant magnitude.
%

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$ steps.
%
If the initial bias potential
before the preliminary run is zero,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%

@@

\subsubsection{\label{sec:mass_distr}
Characterization of optimal schedules}

\begin{table}[h]
  \caption{\label{tab:m_and_a}
    Examples of the variable transformation between the
    reduced schedule, $\alpha^*(t^*)$,
    and the reduced mass distribution, $m^*(\bar q^*)$,
    as defined in Eqs. \eqref{eq:mQ_invTa}
    and \eqref{eq:intmQ_tT}.
  }
  \setlength{\tabcolsep}{2pt} % spacing between columns
  \renewcommand\arraystretch{2.0} % spacing between rows
  \begin{tabular} { c c c }
    \hline
    $m^*(\bar q^*)$ &
    $\alpha^*(t^*)$ &
    $\bar q^*(t^*)$
    \\
    \hline
    $\dfrac{1}{\gamma \, ({\bar q}^* + s_\gamma) }$ &
    $(1 + s_\gamma) \, \gamma \, e^{-\gamma \, t^*}$ &
    $(1 + s_\gamma) \, e^{-\gamma \, t^*} - s_\gamma$
    \\
    $(1+ s_\gamma) \, \gamma \, e^{-\gamma \, {\bar q}^*}$ &
    $\dfrac{1}{ \gamma \, (t^* + s_\gamma) }$ &
    $\dfrac{1}{\gamma} \ln\left(\dfrac{1 + s_\gamma}{t^* + s_\gamma} \right)$
    \\
    $\dfrac{c \, a}{\left( {\bar q}^* + q^*_0 \right)^{a+1}}$ &
    $\dfrac{c^{1/a} \, (1/a)}{\left( t^* + t^*_0 \right)^{(1/a)+1}}$ &
    %$\left( \dfrac{c}{t^* + t^*_0} \right)^{1/a} - q^*_0$
    $q^*_0 \left[\left( \dfrac{1+t^*_0}{t^* + t^*_0} \right)^{1/a} - 1\right]$
    %$\,\begin{aligned}
    %  &a, q^*_0 > 0, \\
    %  &\frac{1}{t^*_0} \equiv \left(1+\frac{1}{q^*_0}\right)^a - 1, \\
    %  &c \equiv {q^*_0}^a(t^*_0 + 1)
    %\end{aligned}$
    \\
    \hline
    \multicolumn{3}{p{8cm}}{
    Here,
    $\gamma > 0$,
    and we have defined
    $s_\gamma \equiv 1/(e^\gamma - 1)$,
    ${t^*_0}^{-1} \equiv \left(1+{q^*_0}^{-1}\right)^a - 1$,
    and
    $c \equiv {q^*_0}^a(1 + t^*_0)$.
    } \\
    \hline
  \end{tabular}
\end{table}

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
We will consider simulations in which
the system is initially equilibrated
at a constant updating magnitude, $a_0$.
%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \mathcal Y\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  =
  \frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible,
which is often true if no eigenvalue is close to $0$,
then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.
}

@@

>>>>

\documentclass[reprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of flat-distribution sampling methods
  for free energy calculation including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the error in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme.
  %
  While the asymptotically optimal schedule for
  the single-bin updating scheme (commonly used in WL)
  is given by an inverse-time formula,
  that for the Gaussian updating scheme (commonly used in metadynamics)
  is often more complex.
  %and the initial updating magnitude is ideally half of
  %the previous equilibrium value.
  %
  Further,
  we show that the single-bin updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence.
  %
  These bandpass updating schemes aim at
  partially flattening a few selected histogram modes
  and their optimal schedule
  is also given by the inverse-time prescription.
  %
  Constructed from orthogonal polynomials,
  bandpass updating schemes serve as generalizations
  of the Langfeld-Lucini-Rago algorithm
  and can be used for parameter tuning for the Gaussian ensemble method
  and umbrella sampling.
  %
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for an adaptive FDS method
of a general updating scheme,
including the single-bin and Gaussian versions.
%
We will map the optimization problem to a mechanical one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
The resulting optimal schedule
depends on the updating window function
through the mass function.
%
While the optimal schedule in the single-bin case
is given by the known inverse-time formula,
that for Gaussian case is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
%In either case, however,
%Ideally,
%the optimal initial updating magnitude
%can be set to
%half of the previous equilibrium value,
%which amounts to a shift of the origin of time
%in the single-bin case.
%
We will also show that
the single-bin updating scheme
is optimal for asymptotic convergence,
and it can be generalized
to a class of asymptotically optimal
bandpass updating schemes.
%
The latter allow one to partially flatten
a few long-range histogram modes,
and to avoid introducing local noise
into the bias potential during the updating process.

The bandpass updating schemes are convenient
in that their optimal schedule is simply given by
the inverse-time formula,
and thus requires no extra calculation.
%
They can be constructed from any orthogonal basis\cite{bussi2006},
and for a polynomial basis,
this construction results in
a class of adaptive FDS algorithms that
update the bias potential based on the
statistical moments of observed distribution
as a generalization of the Langfeld-Lucini-Rago (LLR)
algorithm\cite{langfeld2012, pellegrini2014}.
%
We show that a similar construction
can help FDS algorithms with
quadratic bias potentials\cite{neuhaus2006, *neuhaus2007, zhu2012}
automatically adjust the parameters
to reach a set of Gaussian distributions
with desired means and variances.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we briefly review the basics of FDS
and some known aspects of the optimal schedule.
%
Then, in Sec. \ref{sec:single-bin},
we derive the method of
computing the optimal schedule
in the simple case of the
single-bin updating scheme
used by the WL algorithm,
by proving the optimality
of the known inverse-time formula.
%
Next, we extend the method to the general case
of a multiple-bin updating scheme
that encompasses the Gaussian scheme used in metadynamics
in Sec. \ref{sec:multiple-bin}.
%
We then compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.
%
Finally,
we use the bandpass updating schemes
to build a class of adaptive FDS algorithms
that based on the statistical moments.

@@

From Eqs.
\eqref{eq:vt_diffeq} and \eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  %=
  %\alpha(t) \, \left[ \langle f_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  .
  %\notag
  \label{eq:dxdt_singlebin}
\end{equation}
%
Without the random part, $\zeta_{*i}(t)$,
this equation would drive
the deviation of the bias potential, $v_{*i}(t)$,
to $0$ at long times,
implying the convergence.
%
The formal solution of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot \vartheta\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
\vartheta(q')
&\equiv
e^{q' - q(T)}.
\label{eq:theta_def}
\end{align}

@@

We will first study the eigenvalue decomposition
of the updating matrix, $\mathbf w$,
which helps establish the convergence criterion
and the decomposition of error into
independent eigenmodes.
%
The matrix, $\mathbf w$, is subject to several conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\pmb\rho = (\rho_1, \dots, \rho_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $\rho_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
By a proper overall scaling of $\mathbf w$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will %take advantage of the analogy and
borrow techniques used
in studying transition matrices\cite{vankampen},
but limit ourselves to properties that are unaffected
by the admission of negative elements into $\mathbf w$.

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $x_i$,
%
\begin{equation}
  {\tilde x}_k
  \equiv \mathcal{F}[x_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, x_i
  ,
  %\notag
  \label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%$x_i = \sum_{k = 0}^{n-1} \phi_{ki} \, \tilde{x}_k / \rho_i$.
%
\begin{equation}
  x_i = \frac{1}{\rho_i} \sum_{k = 0}^{n-1} \phi_{ki} \, \tilde{x}_k
  .
  %\notag
  \label{eq:gft_inv}
\end{equation}
%
Then,
we can rewrite Eq. \eqref{eq:ortho0} as
$\mathcal F[1]_k = \delta_{k0}$,
%
and Eq. \eqref{eq:eigenmode0} means that
the bin average
%of any quantity, $x_i$, is the first mode of the Fourier transform:
$\bav{x}= \mathcal F[x_i]_0 = \tilde x_0$,
%
and
\begin{equation}
  \tilde x_{*k}
  = \tilde x_k - \bav{x} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0
    ,
    \\
    \tilde x_k  & k > 0
    .
  \end{dcases}
  \label{eq:xtstar}
\end{equation}
%
Thus, the only effect
of the bin-average deduction on $\tilde x_k$ is to
annihilate the $k=0$ mode.
%
From Eqs. \eqref{eq:gft_def} and \eqref{eq:gft_inv},
we have Parseval's theorem,
\begin{equation}
  \sum_{k=1}^{n-1} \tilde x_k \, \tilde y_k
  =
  \sum_{k=0}^{n-1} \tilde x_{*k} \, \tilde y_{*k}
  =
  \sum_{i=1}^n \rho \, x_{*i} \, y_{*i}
  .
  \label{eq:parseval}
\end{equation}
\note{Proof:
$$
  \sum_{k=0}^{n-1} \tilde x_{*k} \, \tilde y_{*k}
  =
  \sum_{k=0}^{n-1} \tilde x_{*k} \sum_{i=1}^n \phi_{ki} \, y_{*i}
  =
  \sum_{i=1}^n \left( \sum_{k=0}^{n-1} \tilde x_{*k} \, \phi_{ki} \right) \, y_{*i}
  =
  \sum_{i=1}^n \rho \, x_{*i} \, y_{*i}.
$$
}

@@

By Eq. \eqref{eq:parseval},
we can rewrite the error defined in Eq. \eqref{eq:error_def} as
a sum of the errors in individual modes\footnote{If
the error weighting distribution differs from $\rho_i$
in Eq. \eqref{eq:error_def},
the error would contain quadratic cross terms among different eigenmodes.
%
While the solution for the optimal schedule retains the same form,
the mass function defined in Eq. \eqref{eq:mass_func}
will include the corresponding cross terms.
}
%$\Err(T) = \sum_{k = 1}^{n - 1} \left\langle {\tilde v}_k^2(T) \right\rangle$.
%
\begin{align}
  \Err(T)
  =
  \sum_{k = 1}^{n - 1}
    \left\langle
      {\tilde v}_k^2(T)
    \right\rangle
  .
  \notag
\end{align}
%
As in Eq. \eqref{eq:error_tot},
we can decompose the error as the sum
of the residual error and the asymptotic error, with
\begin{align}
  \Err_R(T)
  &=
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) }
  ,
  \label{eq:error_res}
  \\
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot \vartheta_k^2\bigl( q(t) \bigr) \, dt
  ,
  \label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  \vartheta_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\bigl\langle {\tilde \zeta}_k(0)
\, {\tilde \zeta}_k(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty
  \bigl\langle {\tilde \zeta}_k(0)
            \, {\tilde \zeta}_k(t) \bigr\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum} represents
a first approximation for modeling
a general histogram fluctuation, ${\tilde \zeta}_k(t)$,
as an equivalent white noise.
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
In Appendix \ref{sec:Gamma_measure},
we discuss a way of estimating $\Gamma_k$
from an FDS simulation under a constant magnitude.
%

@@

@@

We now determine the optimal value of $q(T)$
that minimizes the total error defined in Eq. \eqref{eq:error_tot}.
%
To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the system undergoes a preliminary FDS simulation
under a constant updating magnitude, $a_0$,
for $T_0$ steps.
%
If the initial bias potential
before the preliminary run is zero,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%

%
From Eqs. \eqref{eq:error_res}, \eqref{eq:xt2_eql1},
and \eqref{eq:error_asym2},
we find that
the optimal $q(T)$ satisfies
%
\begin{align}
  \mathcal Y\bigl( q(T) \bigr)
  &\equiv
  \frac{ M\bigl( q(T) \bigr) } { T }
    \int_0^{q(T)} M(q') \, d q'
  -\frac{a_0}{2} \, M^2\bigl( q(T) \bigr)
  \notag \\
  &
  -\sum_{k = 1}^{n-1}
  \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  =0
  .
\label{eq:opt_qT}
\end{align}
%
%\begin{align}
%  m\bigl( q(T) \bigr)
%  +
%  \frac{2}{a_0 \, C_M \, M\bigl( q(T) \bigr)}
%  \sum_{k = 1}^{n-1}
%  \lambda_k \, \tilde u_k^2 \,
%  e^{-2 \, \lambda_k \, [a_0 \, T_0 + q(T)] }
%  \\=
%  \frac{1} { T \, (a_0 / 2) }
%  .
%\label{eq:opt_qT}
%\end{align}
%
\note{Let $q_T \equiv q(T)$,
$$
\begin{aligned}
  \frac{
    \partial \Err_R(T)
  }
  {
    \partial q_T
  }
  &=
  -a_0 \, M^2(q_T)
  -2
  \sum_{k=1}^{n-1} \lambda_k \,
  \tilde u_k^2 e^{-2 \, \lambda_k \, (a_0 \, T_0 + q_T)}
  ,
  \\
  \frac{
    \partial \Err_A(T)
  }
  {
    \partial q_T
  }
  &=
  \frac 2 T \,
  M(q_T) \,
  \int_0^{ q_T } M(q) \, dq
  =
  \frac{ 2 \, C_M } { T } \, M(q_T)
  .
\end{aligned}
$$
Then solve $\partial (\Err_R + \Err_A) / \partial Q_c = 0$.
}
%
We can rewrite the condition in terms of
the initial updating magnitude by Eq. \eqref{eq:Lagrangian_const} %\eqref{eq:mQ_invTa},
\begin{align}
  \alpha(0)
  %=
  %\frac{1}{T \, m\bigl( q(T) \bigr) }
  =
  \frac{ a_0 } { 2 }
  +
  \sum_{k = 1}^{n-1}
  \frac{
    \lambda_k \, \epsilon_k \, e^{-2\lambda_k \, q(T)}
  }{M^2\bigl( q(T) \bigr)}
  .
  \label{eq:opt_alpha0}
\end{align}
%
If the initial bias, $\epsilon_k$, is negligible,
which is often true if no eigenvalue is close to $0$,
then
%
\begin{equation}
  \alpha( t = 0 )
  \approx
  \frac{ a_0 }
       { 2 }
  ,
\label{eq:half_alpha0}
\end{equation}
%
i.e. the optimal initial updating magnitude
is half of the equilibrium value.\footnote{This factor, $1/2$,
happens to coincide with the
recommended reduction factor of the updating magnitude
at stage transitions
in the original WL algorithm\cite{
wang2001, wang2001pre}.
}

@@

\section{\label{sec:schedule_geometry}
Characterization of optimal schedules}
